\chapter{The MMIX Architecture}
\label{ch:mmix-arch}

As already mentioned in the introduction, MMIX is a 64-bit \glslink{Endianness}{big-endian} \gls{RISC} machine. It provides 256 32-bit wide instructions, at least 256 general purpose registers and 32 special registers. Both are 64-bit wide. Additionally it has a 64-bit virtual and physical address space and supports both integer arithmetic and floating point arithmetic. \gls{Donald Knuth} described the goals of MMIX with "I strove to design MMIX so that its machine language would be simple, elegant, and easy to learn. At the same time I was careful to include all of the complexities needed to achieve high performance in practice, so that MMIX could in principle be built and even perhaps be competitive with some of the fastest general-purpose computers in the marketplace." \citep[pg. v]{mmix-ware-book}.

This chapter splits the features of MMIX in categories and describes them one after another. In each category the concept is explained, if necessary, and the associated instructions are introduced. It will mostly resemble the MMIX specification \citep{mmix-doc}, but of course, this thesis has a different purpose, \ie some parts will be described in more detail, some in less. Especially, this thesis will try to give more examples to the difficult chapters of MMIX. But after all, it is of course not meant to be a replacement for the specification. Thus, whenever a concept or instruction is explained, the end of it will link to the corresponding page of the specification.

\input{mmix-arch-intro}
\input{mmix-arch-reg}
\input{mmix-arch-intarith}
\input{mmix-arch-bit}
\input{mmix-arch-cmp}
\input{mmix-arch-branch}
\input{mmix-arch-mem}
\input{mmix-arch-float}
\input{mmix-arch-regstack}
\input{mmix-arch-intrptex}

